Bayesian regression facilitates quantitative modelling of cell metabolism

### 2023-10-31 paper/main.qmd
Paper update

The parameters must include sufficient boundary concentrations and fluxes to
solve \eqref{eq-1}.

@@

In a fermentation context, \eqref{eq-3} captures the rapid kinetics inside
the cell, while another set of ODEs would be used to describe the external
substrate and product concentrations, which could act as boundary parameters
to \eqref{eq-3}.

@@

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy models in which measurable
quantities are sensitive to combinations of parameters but not to individual
marginal parameter values are ubiquitous in models of biological systems
[@gutenkunst_2007; @white_2016]. The parameter correlation structure represents
the set of potential models that describe the observed data. As we demonstrate
in our case study below, capturing this correlation structure is difficult
outside a Bayesian context.

@@

There have also been efforts to implement Bayesian inference for kinetic models
without the use of MCMC. Examples of alternative inference methods include
variational inference as in @st.johnBayesianInferenceMetabolic2018, rejection
sampling and approximate Bayesian computation @saa_construction_2016 and Laplace
approximation, in which the Fisher information matrix is used to calculate a
normal approximation around the maximum a posteriori parameter configuration
[@liebermeister_model_2021; @gopalakrishnan_k-fit_2020; @stapor_pesto_2018;
@raue_joining_2013]. Non-MCMC-based Bayesian kinetic models have limited
utility because they lack reliable diagnostic tools for verifying that their
results approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian, joint
probability distributions [@gutenkunst_2007; @stapor_pesto_2018].

Our application Maud is the first Bayesian kinetic model to combine biologically
realistic mechanistic accuracy---including accurate rate laws, post-
translational modification and thermodynamics---with fast, robust MCMC sampling
using adaptive Hamiltonian Monte Carlo. Further, Maud is a general-purpose
application that can be used to fit a wide range of Bayesian kinetic models.

@@

To demonstrate our application’s capabilities, we used Maud to analyse an
artificial dataset based on the human methionine cycle. We first used Maud
to generate simulated training and validation measurements based on plausible
parameter values, then performed posterior sampling. Next, we used Maud to
predict the validation data.

To show that Maud is robust to missing measurements we compared the results
of fitting the full dataset with an intentionally incomplete dataset. To
demonstrate why a full Bayesian approach is preferable to an approach based on a
Laplace approximation of the posterior distribution, we also fit our methionine
cycle model using the latter method and compared the results with MCMC sampling.

Finally, we investigated our results to find out what our model learned about
the contributions of different regulatory factors to the flux through GNMT, an
important reaction. This analysis illustrates how Maud can be used to generate
actionable insights about metabolism without the need for further statistical
analysis.

>>>>

The parameters must include sufficient boundary concentrations and fluxes to
solve equation \eqref{eq-1}.

@@

In a fermentation context, equation \eqref{eq-3} captures the rapid kinetics
inside the cell, while another set of ODEs would be used to describe the
external substrate and product concentrations, which could act as boundary
parameters to equation \eqref{eq-3}.

@@

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy models in which measurable
quantities are sensitive to combinations of parameters but not to individual
marginal parameter values are ubiquitous in models of biological systems
[@gutenkunst_2007; @white_2016]. The parameter correlation structure represents
the set of potential models that describe the observed data. As we demonstrate
in @sec-laplace below, capturing this correlation structure is difficult outside
a fully Bayesian context.

@@

There have also been efforts to implement Bayesian inference for kinetic models
without the use of MCMC. Examples of alternative inference methods include
variational inference [@st.johnBayesianInferenceMetabolic2018], rejection
sampling and approximate Bayesian computation [@saa_construction_2016] and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori parameter
configuration [@liebermeister_model_2021; @gopalakrishnan_k-fit_2020;
@stapor_pesto_2018; @raue_joining_2013]. Non-MCMC-based Bayesian kinetic
models have limited utility because they lack reliable diagnostic tools for
verifying that their results approximate the target posterior distribution.
This is a problem because realistic kinetic models tend to induce highly
correlated, non-Gaussian, joint probability distributions [@gutenkunst_2007;
@stapor_pesto_2018].

Our application Maud provides the first Bayesian kinetic model to combine
biologically realistic mechanistic accuracy---including accurate rate laws,
post- translational modification and thermodynamics---with fast, robust
MCMC sampling using adaptive Hamiltonian Monte Carlo. Further, Maud is a
general-purpose application that can be used to fit a wide range of Bayesian
kinetic models.

@@

To demonstrate our application’s capabilities, we used Maud to analyse an
artificial dataset based on the human methionine cycle. We generated this
dataset using Maud, by simulating measurements for a set of training and
validation conditions based on plausible parameter values. Next, we performed
posterior inference for the training measurements and prediction for both
training and validation measurements.

We investigated Maud's sensitivity to missing measurements by comparing the
results of fitting a full dataset with an intentionally incomplete dataset. To
demonstrate why a full Bayesian approach is preferable to an approach based on a
Laplace approximation of the posterior distribution, we compared the results of
analysing a representative metabolic network using both methods.

Finally, we dug into our results to find out what our full dataset methionine
model learned about the contributions of different regulatory factors to the
flux through GNMT, an important reaction. This analysis illustrates how Maud can
be used to generate actionable insights about metabolism without the need for
further statistical analysis.
